\documentclass[10pt]{article}
\setlength{\oddsidemargin}{.5in}
\setlength{\evensidemargin}{.5in}
\setlength{\textwidth}{6.0in}
\setlength{\topmargin}{0.0in}
\addtolength{\topmargin}{-1.0in}
\setlength{\footskip}{.5in}
\setlength{\textheight}{9.5in}
\linespread{1.1}

\usepackage[dvips]{graphicx}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mjcquiz}

% Michael Carlisle, Baruch College, CUNY, MTH 4100, Spring 2018
% github.com/mcarlisle/course-materials (posted Jan 2019)

\newcommand{\exam}[2]{
\namefield

\headerexam{Linear Algebra}{Exam \#2}

\instexam{100}{105}

\pagebreak

%\headerexam{Linear Algebra}{Exam \#2}

%Work for Problem 1 here

%\pagebreak

\headerexam{Linear Algebra}{Exam \#2}

\prob{1} (10+5+10+5 pts) Consider the system $A \vec{x} = b$ given by  % 3 x 5 system
#1
\begin{itemize}
\item[(a) ] What is the reduced row echelon form of the coefficient matrix $A$? 
\item[(b) ] What are the pivot variables? What are the free variables?
\item[(c) ] What is the full (particular + special) solution set to the system $A \vec{x} = b$? %Give in terms of the special solution(s). 
\item[(d) ] What is the dimension of the solution set $\{ \vec{x} : \,\, A \vec{x} = b\}$? \\
What geometric shape does this set have, inside what larger vector space? 
\end{itemize}

\pagebreak

%\headerexam{Linear Algebra}{Exam \#2}

%Work for Problem 2 here

%\pagebreak

\headerexam{Linear Algebra}{Exam \#2}

\prob{2} (20+5 pts) A line in $\mathbb{R}^3$ is spanned by the first column of $A$ from Problem 1. 
\begin{itemize}
\item[(a) ] Give the projection matrix $P$ that projects a vector in $\mathbb{R}^3$ onto this line.
\item[(b) ] What is the projection of $\cvvv{1}{0}{0}$ onto this line?
\end{itemize}

\pagebreak

%\headerexam{Linear Algebra}{Exam \#2}

%Work for Problem 3 here

%\pagebreak

\headerexam{Linear Algebra}{Exam \#2}

\prob{3} (20+5+5 pts) 
\begin{itemize}
\item[(a) ] Find the best fit quadratic function $f(x) = a + bx + cx^2$ to the set of points 
\[ \{ #2 \}. \]
\item[(b) ] Plot the graph and points.
\item[(c) ] Compute the total squared error from the best fit parabola to the points. 
\end{itemize}

\pagebreak

%\headerexam{Linear Algebra}{Exam \#2}

%Work for Problem 4 here

%\pagebreak

\headerexam{Linear Algebra}{Exam \#2}

\prob{4} (20 pts) Compute the Gram-Schmidt basis for $\mathbb{R}^3$ generated by the independent columns of $A$ based on the pivot variables you found during Problem 1. 

(Yes, you read that right. If you read this first, before starting Problem 1, congratulations, you get a big hint on Problem 1.)

Remember: 
\begin{align*} 
b_1 = a_1 & \implies q_1 = \frac{1}{||b_1||} b_1 \\
b_2 = a_2 - \left( \frac{b_1 \cdot a_2}{b_1 \cdot b_1} \right) b_1 & \implies q_2 =  \frac{1}{||b_2||} b_2...
\end{align*}

}


%---:----1----:----2----:----3----:----4----:----5----:----6----:----7----:---

\begin{document}

\pagenumbering{arabic}

\exam{\begin{align*}
4 x_1 + 21 x_2 + 97 x_3 - 12 x_4 + 19 x_5 & = 111 \\
8 x_1 + 35 x_2 + 159 x_3 - 9 x_4 + 45 x_5 & = 171 \\
2 x_1 + 7 x_2 + 31 x_3 + 3 x_4 + 13 x_5 & = 27.
\end{align*}}{(0,5), (-2,8), (2,9), (1,6)}


\end{document}
