\documentclass[10pt]{article}
\setlength{\oddsidemargin}{.5in}
\setlength{\evensidemargin}{.5in}
\setlength{\textwidth}{6.0in}
\setlength{\topmargin}{0.0in}
\addtolength{\topmargin}{-1.0in}
\setlength{\footskip}{.5in}
\setlength{\textheight}{9.5in}
\linespread{1.1}

% Michael Carlisle, Baruch College, CUNY, MTH 2610, Fall 2016
% github.com/mcarlisle/lecture-notes (posted Jan 2019)

\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}


\begin{document}

\pagenumbering{gobble}

%---:----1----:----2----:----3----:----4----:----5----:----6----:----7----:---

\begin{center}
Calculus I Quiz \#8 (sample final exam; use your own paper)
\end{center}


\noindent\textbf{Problem 1} \,\, Assume $g(x)$ is a differentiable and integrable function over the real numbers. Let 
%, and that $G(x) = \int g(x) dx$. Let 
\[ f(x) = 9e^{-2x} + 6 \sin( g(x) ) - \frac{3}{9x+4}.\]

\begin{itemize}
\item[a) ] What is the domain of $y=f(x)$?
\item[b) ] Find $f'(x)$ and $f''(x)$.
\item[c) ] Find $\int f'(x) dx$.
\item[d) ] Find $\int_{1}^{10} f'(x) dx$. (No need for a calculator; if there are $e$ or $\ln$ or $\sin$ in your answer, leave them as they are. Just simplify the numbers as best you can.)
\item[e) ] Let $g(x) = 3x + 2$. Find $\int_{3}^{6} f(x) dx$.
\end{itemize}

\vspace{5mm}

\noindent\textbf{Problem 2} \,\, Solve the differential equation $y' = -5y$ for $y = f(x)$ subject to the initial condition $f(0) = 12$.

\vspace{5mm}

\noindent\textbf{Problem 3} \,\, Assuming the acceleration due to gravity is $-9.8 \frac{\text{m}}{\text{sec}^2}$ on Earth, how high was an object dropped from rest if it took exactly 10 seconds to reach the ground?

\vspace{5mm}

\noindent\textbf{Problem 4} \,\, Air enters a (spherical) balloon at the rate of 5 cubic feet per minute ($\frac{\text{ft}^3}{\text{min}}$). Recalling that the volume of a sphere is $V = \frac{4}{3} \pi r^3$, where $r$ is the radius of the sphere (here, in feet), what is the rate of the radius' growth when the balloon's volume is 400 $\text{ft}^3$? (This is a large balloon.)

\vspace{5mm}

\noindent\textbf{Problem 5} \,\, Minimize the average cost per item of a product with total cost function 
\[ C(x) = (100 + 5e^{-2x})x + 0.1x^2. \]

\vspace{5mm}

\noindent\textbf{Problem 6} \,\, Let $f(x) = x h(x^2)$, and $\int h(x) dx = H(x) + C$. What is $\int f(x) dx$?

\vspace{5mm}

\noindent\textbf{Problem 7} \,\, Find $\int_a^b (c x^n + \frac{5}{x}) dx$ for $n=-1$ and $n \neq -1$.

\vspace{5mm}

\noindent\textbf{Problem 8} \,\, 
\begin{itemize}
\item[(a) ] Approximate $\int_{0}^{4} (x^2 + 6) dx$ by calculating a sum, with right-hand endpoints, for equal box spacing size $\Delta x = 1$ (that is, 4 boxes to check area). Sketch the graph to help you along. 
\item[(b) ] Calculate the actual integral $\int_{0}^{4} (x^2 + 6) dx$. How far off was your approximation?
\end{itemize}

\vspace{5mm}

\noindent\textbf{Problem 9} \,\, Compute $\frac{d}{dx} \tan x$ via the quotient rule (you must show your work).

\vspace{5mm}

\noindent\textbf{Problem 10} \,\, Find $\int \tan(x) dx$ (you must show the substitution). 

\vspace{5mm}

\noindent\textbf{Problem 11} \,\, Calculate:
\begin{itemize}
\item[(a) ] $\frac{d}{dx} \log_{3}(x)$
\item[(b) ] $\int 3^x dx$
\item[(c) ] $4^{\log_2 x}$
\end{itemize}


\end{document}
