\documentclass[10pt]{article}
%\setlength{\oddsidemargin}{.5in}
%\setlength{\evensidemargin}{.5in}
\setlength{\textwidth}{5.5in}
\setlength{\topmargin}{0.0in}
\addtolength{\topmargin}{-0.5in}
\setlength{\footskip}{.5in}
\setlength{\textheight}{9.4in}
\linespread{1.1}

% Michael Carlisle, Baruch College, CUNY, MTH 3010, Fall 2017
% github.com/mcarlisle/course-materials (posted Jan 2019)

\usepackage[dvips]{graphicx}
\usepackage{epsfig}
\usepackage{amssymb}
\usepackage{amsmath}

\newcommand{\prob}[1]{\vspace{10mm} \noindent \textbf{Problem #1} \,\,}
\newcommand{\header}{
\begin{center}
Calculus II Quiz \#2
\end{center}

\vspace{2mm}

}

\newcommand{\namefield}{
\noindent Printed Name \underline{\hspace{50mm}} \,\,\, Signature \underline{\hspace{50mm}}
}



\newcommand{\inst}[2]{
Show your work and clearly label your answers on this quiz. \emph{No scrap paper, calculators, or notes are allowed} (or needed). This quiz is scored out of #1 points. (There are #2 points possible.) You have 30 minutes to complete the quiz.

To get credit on a problem, you \emph{must} show work. Even if you can do the work in your head, the point of these exercises is to get you to articulate your thought processes.
}


\begin{document}

\pagenumbering{arabic}

%---:----1----:----2----:----3----:----4----:----5----:----6----:----7----:---

\namefield

\header

\inst{40}{50}


\prob{1} (5+10 pts) % 7.1

\begin{itemize}
\item[(a) ] Sketch the graph of the region completely enclosed by the curves 
\begin{align*}
f(x) & = 4 - x^2 \text{ and } g(x) = x^2 + 2x.
\end{align*}
\item[(b) ] Compute the area of the region sketched in (a).
\end{itemize}

\pagebreak

\header

\prob{2} (5+10+10 pts) % 7.2, 7.3

\begin{itemize}
\item[(a) ] Sketch the graph of the region contained by the curves 
\[ y = \frac{2}{x}, \,\, x = 1, \,\, x=2, \,\, y = 0. \] 
\item[(b) ] Compute the volume of the solid obtained by revolving the curve in (a) around the $y$-axis \emph{using the washer method}. Explain all your steps.
\item[(c) ] Compute the volume of the solid obtained by revolving the curve in (a) around the $y$-axis \emph{using the shell method}. Explain all your steps.
\end{itemize}

\pagebreak

\header

\prob{3} (10 pts) % 7.4

What is the arc length of the curve $y = \sqrt{1 - x^2}$ over the interval $x \in [-1,1]$?


\end{document}
